Numerical simulation of two-dimensional heat convection-diffusion with ellipsoid geometries

### 2016-02-01 article/main.tex
Closer to the final draft

\usepackage{amsmath, amssymb, amsthm}

@@

\begin{abstract}
    Ever wonder what's the how to keep your warm bath at the right temperature
    while helping the environment? This paper introduces and analyzes
    two-dimensional heat convection-diffusion through the Navier-Stokes
    formulation of fluid-flow and the classical heat equation. Our mission is to
    determine the optimal amount of water necessary to maintain the perfect
    bathtub temperature. BLAH BLAH BLAH
\end{abstract}

@@

Figure 1: The heat diffusion process in the case of a stationary fluid and
constant faucet tap. The mesh
dimensions are $50 \times 50$, $\alpha = 10^{-3}$, $T_{\mathrm{tub}} = 10$, and
$T_{\mathrm{faucet}} = 30$.
\end{center}

@@

    Figure 2: Temperature total variation over time in the case of a stationary
    fluid, with the same parameters as in Figure 1. The monotonicity of the TV
    function is in line with basic intution.
\end{center}

@@

    Figure 3: The heat diffusion process in the case of a stationary fluid and
    temporary faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$.
\end{center}

@@

    Figure 4: Temperature total variation over time in the case of a stationary
    fluid, with the same parameters as in Figure 3. The monotonicity of the TV
    function is in line with basic intution.
\end{center}
The fact that the total variation performs so poorly in this case has
implications for the policy problem, which we will discuss in the next section.

Generalizing beyond stationary fluids reveals a number of critical issues of the
numerical techniques employed in this study. In stationary fluid, the
Navier-Stokes and Poisson equations are not in play; however, by relaxing the
stationarity constraint, we observe the effects of the numerical instability of
our approximations to these equtaions. We simulate the case described in Figure
1 but relax the stationarity constraint.

@@

    Figure 5: The heat diffusion process in the case of a moving fluid and
    temporary faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$. The
    parameter controlling the fluid motion is set $b = 5$.
\end{center}

By comparing Figure 5 with Figure 1, we see that adding the convection process
into the model actually improves the spread of the faucet temperature. Since
more of the bathtub area is white in the images, we have more of the surface
area heated by the temperature water.

@@

    Figure 6: Temperature total variation (with respect to $T_{\mathrm{faucet}}$
    and with respect to the mean bathtub temperature over time in the case of a
    nonstationary fluid, with the same parameters as in Figure 5. The TV
    functions are considerably less well behaved, but notably the TV with
    respect to $T_{\mathrm{faucet}}$ decreases to a lower level than the same TV
    in Figure 2.
\end{center}

@@

    Figure 7: The heat diffusion process in the case of a moving fluid and
    temporary faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$. The
    parameter controlling the fluid motion is set $b = 8$. The numerical
    instability of the simulation causes the complete whitewashing by the final
    frame.
\end{center}

By comparing Figure 5 with Figure 1, we see that adding the convection process
into the model actually improves the spread of the faucet temperature. Since
more of the bathtub area is white in the images, we have more of the surface
area heated by the temperature water.

@@

    Figure 8: Temperature total variation (with respect to $T_{\mathrm{faucet}}$
    and with respect to the mean bathtub temperature over time in the case of a
    nonstationary fluid, with the same parameters as in Figure 5. The TV
    functions are considerably less well behaved, but notably the TV with
    respect to $T_{\mathrm{faucet}}$ decreases to a lower level than the same TV
    in Figure 2.
\end{center}

The numerical instability persists even when the faucet is turned off, as in
Figure 3.

@@

As we showed in the previous section, our current numerical implementation of
the Navier-Stokes equations are highly sensitive to the initial conditions of
the model. Though a careful calibration of the deep parameters (e.g. $\rho$,
$\alpha$, $\beta$, etc.) and mesh granularity might produce more stable results,
we doubt that this would be sufficient. Fundamentally, forward Euler
approximations of the solutions to differential equations encourage volatility 
by accumulating the relatively small errors from the individual step
calculations into catastrophically unstable general results. To prevent this
kind of instability requires a re-examination of the numerical approach to
approximating these PDEs.

@@

That maintaining an active source (in the case of the bathtub narrative, a
running faucet) reduces the total variation of the temperature in the bathtub
supports our formulation of the optimization problem and also indicates suitable
choices for $\gamma$. If the optimization problem requires a bound

>>>>

\usepackage{amsmath, amssymb, amsthm}
\usepackage{ragged2e}

@@

\begin{abstract}
\end{abstract}

@@

\justify
\footnotesize{
Figure 1: The heat diffusion process in the case of a stationary fluid and
constant faucet tap. The mesh
dimensions are $50 \times 50$, $\alpha = 10^{-3}$, $T_{\mathrm{tub}} = 10$, and
$T_{\mathrm{faucet}} = 30$.}
\end{center}

@@

    \justify
    \footnotesize{
    Figure 2: Temperature total variation over time in the case of a stationary
    fluid, with the same parameters as in Figure 1. The monotonicity of the TV
function is in line with basic intution.}
\end{center}

@@

    \justify
    \footnotesize{
    Figure 3: The heat diffusion process in the case of a stationary fluid and
    temporary faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$.}
\end{center}

@@

    \justify
    \footnotesize{
    Figure 4: Temperature total variation over time in the case of a stationary
    fluid, with the same parameters as in Figure 3. The monotonicity of the TV
function is in line with basic intution.}
\end{center}
The fact that the total variation performs so poorly in this case has
implications for the policy problem, which we will discuss in the next section.

Generalizing beyond stationary fluids reveals a number of critical issues of the
numerical techniques employed in this study, but also provides useful
information regarding the underlying problem. In stationary fluid, the
Navier-Stokes and Poisson equations are not in play; however, by relaxing the
stationarity constraint, we observe the effects fluid flow on the diffusivity of
heat. We simulate the case described in Figure 1 but relax the stationarity constraint.

@@

    \justify
    \footnotesize{
    Figure 5: The heat diffusion process in the case of a moving fluid and
    permanent faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$. The
parameter controlling the fluid motion is set $b = 5$.}
\end{center}

By comparing Figure 5 with Figure 1, we see that adding the convection process
into the model actually improves the spread of the faucet temperature. Since
more of the bathtub area is white in the images, we have more of the surface
area heated by the temperature water. This phenomenon is confirmed by measuring
the total variation of the bathtub water over time.

@@

    \justify
    \footnotesize{
    Figure 6: Temperature total variation (with respect to $T_{\mathrm{faucet}}$
    and with respect to the mean bathtub temperature over time in the case of a
    nonstationary fluid, with the same parameters as in Figure 5. The TV
    functions are considerably less well behaved, but notably the TV with
    respect to $T_{\mathrm{faucet}}$ decreases to a lower level than the same TV
in Figure 2.}
\end{center}

There are three interesting phenomena present in this result. First, we see that
the evolution of the total variation over time is less well-behaved than before.
We also note that the total variation with respect to the mean bathtub
temperature grows almost linearly with time and exhibits much less concavity
than Figure 2. Finally, the total variation with respect to
$T_{\mathrm{faucet}}$ reaches a lower threshold than the total variation in the
diffusion case only in Figure 2.

The choice of $b=5$ is essntially arbitrary; however the behavior of the fluid
with respect to different choices is highly volatile. In fact, modest changes to
the value for $b$ have drastic implications for the numerical stability of the
fluid flow, and consequently the temperature. We see this phenomenon in Figures
7 and 8.

@@

    \justify
    \footnotesize{
    Figure 7: The heat diffusion process in the case of a moving fluid and
    permanent faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$. The
    parameter controlling the fluid motion is set $b = 8$. The numerical
    instability of the simulation causes the complete whitewashing by the final
frame.}
\end{center}

The evolution of frames in Figure 7 represents an explosion in the numerical
values associated with the temperatures. The black-white ridge in the bottom
right corner of the third frame exhibits the typical instability phenomenon we
observed in the simulations. Emanating from near the faucet source, this
phenomenon appears to push out towards the boundary of the frame until the
entire frame is whitewashed. We can also see this last-minute explosion by
measuring the total variation.

@@

    \justify
    \footnotesize{
    Figure 8: Temperature total variation (with respect to $T_{\mathrm{faucet}}$
    and with respect to the mean bathtub temperature over time in the case of a
    nonstationary fluid, with the same parameters as in Figure 5. The TV
    functions are considerably less well behaved, but notably the TV with
    respect to $T_{\mathrm{faucet}}$ decreases to a lower level than the same TV
in Figure 2.}
\end{center}

The presence of a constant heat and flow source in the faucet might suggest that
the numerical instability associated with the parametrization of $b$ relates
also to the source. We see that this is not the case. The numerical instability
persists even when the faucet is turned off, as in Figure 3.

\begin{center}
    \includegraphics[width=0.8\textwidth]{../plots/diffusion-05.png}

    \justify
    \footnotesize{Figure 9: The heat diffusion process in the case of a moving fluid and
    temporary faucet tap. The mesh dimensions are $50 \times 50$, $\alpha =
    10^{-3}$, $T_{\mathrm{tub}} = 10$, and $T_{\mathrm{faucet}} = 30$. The
    parameter controlling the fluid motion is set $b = 8$. The numerical
    instability of the simulation causes the complete whitewashing by the final
frame.}
\end{center}

Interestingly, the second frame seems to show numerical instability towards
underflow, but by the third frame we observe near complete whitewashing of the
same characterization as Figure 7. We also see the black-white ridge in the
bottom right corner of the third frame, again in a similar fashion to Figure 7.

More surprisingly is the behavior of the temperature total variation, as we see
below in Figure 10.
\begin{center}
    \includegraphics[width=0.5\textwidth]{../plots/tv-05.png}

    \justify
    \footnotesize{
        Figure 10: Temperature total variation (with respect to
        $T_{\mathrm{faucet}}$ and with respect to the mean bathtub temperature
        over time in the case of a nonstationary fluid, with the same parameters
        as in Figure 9. The TV functions are very poorly behaved and exhibit no
        similar characteristics to the previous configurations of the
        simulation.}
\end{center}
We have no intuition or explanation for the behavior of these total variation
functions. We suspect that this is almost entirely due to numerical artifacts in
the simulation but cannot speak with any authority to why the hump shape appears
in the plots. We do note, however, that the hump appears to correspond to the
second frame of Figure 9, in which we observe a very dark region with low
temperature.

@@

As we showed in the previous section, our current numerical implementation of
the Navier-Stokes equations are highly sensitive to the initial conditions of
the model. Though a careful calibration of the deep parameters (e.g. $\rho$,
$\alpha$, $b$, etc.) and mesh granularity might produce more stable results,
we doubt that this would be sufficient. Fundamentally, forward Euler
approximations of the solutions to differential equations encourage volatility 
by accumulating the relatively small errors from the individual step
calculations into catastrophically unstable general results. To prevent this
kind of instability requires a re-examination of the numerical approach to
approximating these PDEs.

@@

That maintaining an active source (in the case of the bathtub narrative, a
running faucet) reduces the total variation of the temperature in the bathtub
supports our formulation of the optimization problem and also indicates suitable
choices for $\gamma$. Since including nonstationarity reduces the total
variation of the temperature the farthest, we can suggest to the avid bather
that keeping the water moving in the bathtub likely keeps the bathtub
temperature more evenly distributed than letting the water approach
stationarity.

Further investigation should attempt to mitigate the detrimental effects of
unchecked numerical instability. The mesh approach to the numerical
approximation of PDE solutions can be optimized \cite{betounes, 12-steps} to
minimize the accumulation of error. Solving our proposed optimization problem
requires additional study, too. Ultimately, our avid bather needs a policy to
maximize her utility, so specifying the ideal velocity $c$ remains the future
priority.
